---
title: "Clinical Data Analysis Notes"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---



<details> <summary> About the author 😃 Chris Battiston </summary>
I am the Research Database Administrator for a hospital in Toronto, Canada.  I have over 20 years experience in database administration and all things healthcare data-related.  I have been using R now for over 7 years, and have become increasingly appreciative of it's flexibility, power, and user-friendly environment.  


I am self-taught for many of these skills, so you can do it too!

I'm a hardcore fan of heavy metal music, coffee, and ridiculously spicy food🔥, and spend my spare time chasing my 5 1/2 year old son around the house (or cleaning up his never-ending piles of PlayDoh, Legos, and Hot Wheels🚗!).

I'm a Geek that loves data, R, and passionate about taking complex topics and breaking them down for easier consumption.  
</details>

<br>

This is the more detailed version of the presentation given for the [ITN Workshop at the University of Michigan on Analyzing Clinical Data](https://hutchdatascience.org/UMich_ITN_Workshop/clinical_data.html).  


❓ Questions about the content (but not support questions!) can be emailed to chris.battiston@wchospital.ca.❓ 

The goals are to:

- Give an introduction to R and tools for clinical analysis
- Give a more thorough introduction to these tools:
  - `sqlr` - Your best friend in data management
  - `ggplot2` - Your best friend for visualizations
  - `survival` - Your best friend for analyzing survival data
  - `survminer` - Your best friend for making publication ready survival analysis data plots
- Use a sample dataset to dig into survival or Kaplan-Meier curves

## Clinical Data Analysis Considerations

Clinical data analysis is a complex process that requires careful attention to the following to ensure accurate and meaningful results:

- data integrity
- regulatory compliance
- security
- statistical rigor 

High-quality data is the foundation of reliable analysis. Data cleaning, standardization, and validation are essential first steps: 

- **Cleaning** involves handling missing values, resolving duplicates, and correcting inconsistencies
- **Standardization** ensures uniformity through common data formats such as CDISC, SDTM, or ADaM. 
- **Validation** techniques, including range checks, logical consistency checks, and audit trails, further enhance data reliability.

Compliance with regulatory and ethical guidelines is paramount in clinical data analysis. Regulations such as HIPAA, PHIPA, and GDPR dictate stringent rules on data privacy and security. Ethical considerations also play a crucial role, requiring approval from an Institutional Review Board (IRB) or Research Ethics Board (REB) before conducting research involving human data. De-identification techniques, including data masking and anonymization, must be implemented when sharing or analyzing patient data to protect personal information and ensure compliance with ethical and legal standards.

Data security and controlled access are critical in clinical research, given the sensitivity of health-related information. The following practices can help ensure the data remains secure:

- **Role-based access control (RBAC)** ensures that only authorized personnel can view or modify specific data, reducing the risk of breaches. 
- **Encryption techniques** must be applied to protect data both at rest and in transit, ensuring that patient information remains secure. 
- **Audit logs** of data access and modifications supports accountability and helps identify potential security threats.

The choice of analytical methods depends on the study objectives and the nature of the dataset. Descriptive statistics, such as means, medians, and distributions, provide insights into patient demographics and baseline characteristics. Inferential statistics, including hypothesis testing, regression analysis, and survival analysis, help in understanding relationships between variables and making predictions. Increasingly, machine learning models are being applied in clinical research for predictive analytics, risk stratification, and pattern detection. However, such models must be interpreted with caution, ensuring transparency, bias mitigation, and clinical relevance.

Reproducibility and transparency are crucial for ensuring the credibility of clinical data analyses. Maintaining well-documented protocols, version-controlled scripts, and metadata helps facilitate data traceability and consistency across studies. Open-source platforms, such as GitHub or Jupyter Notebooks, enable researchers to share code and methodologies, promoting collaboration and validation. Standard Operating Procedures (SOPs) should be established to outline clear and repeatable methodologies for data processing, ensuring consistency across different analyses and research teams.

Integration with external systems further enhances the efficiency and scalability of clinical data management. The following can help:

- **Compatibility with Electronic Health Records (EHR)** and other health information systems enables seamless data exchange. 
- **Application Programming Interfaces (APIs)** and automated data pipelines streamline data extraction, transformation, and loading (ETL) processes, reducing manual errors and improving efficiency. 
- **Data linkage techniques**, using unique patient identifiers, allow researchers to combine information from multiple sources, leading to more comprehensive insights.

Finally, researchers must be mindful of potential biases and the interpretation of findings. Confounding variables should be identified and adjusted for in study designs to avoid misleading conclusions. Additionally, ensuring that study findings are generalizable beyond the sample population is crucial for clinical applicability. Emphasizing clinical relevance rather than mere statistical significance is essential, as results should ultimately contribute to improved patient care, treatment decisions, and healthcare policies.


## Introduction to R

R is a massive, fully customizable data management environment.  To do it any justice, I'd need a week of classes - so given this is based on a short workshop I gave, I'm going to cover a lot of content that I was not able to cover in the session. 


R is driven largely by what are called "libraries" or "packages", which are basically plugins or extensions created by the R User Community.  For every one task you can do in R, there is likely going to be a multitude of packages you can choose from - as you move forward in becoming an R User, spend time familiarizing yourself with the various packages to see what each offers. 

Designed for statistical computing, data analysis, and visualization, R is widely used in academia, research, and industry for tasks ranging from basic data manipulation to advanced machine learning. As mentioned, one of R’s key strengths is its extensive package ecosystem, which allows users to extend its functionality by installing and calling specialized libraries such as `ggplot2` for visualization, `dplyr` for data manipulation, and `survival` for survival analysis. Installing and loading libraries is crucial because many of R’s advanced features are provided through community-contributed packages rather than built into the base language.

### R vs SAS 

Compared to SAS, R offers greater flexibility, a more active community, and as it's open-source, R has better support for cutting-edge statistical techniques and machine learning. While SAS is known for its reliability in enterprise environments, R’s extensive visualization capabilities and its integration with modern tools like Shiny and R Markdown make it a more versatile choice. R is also free, whereas SAS requires a costly license, making R more accessible to individuals, startups, and educational institutions.

R’s increasing popularity is driven by the rise of data science, the availability of extensive online learning resources, and its ability to integrate with other programming languages like Python and SQL. It is particularly favored for research and academic work because of its strong statistical modeling capabilities and the transparency of its open-source development. Additionally, companies are adopting R for business analytics, finance, healthcare, and bioinformatics, further solidifying its role in modern data-driven decision-making.



## Clinical Data Tools Beyond R

Beyond R, effective clinical data management and analysis require specialized tools that facilitate data collection, storage, security, and compliance with regulatory standards. Several platforms are widely used in clinical research, each offering unique functionalities to support various study designs and workflows. Among these, OpenClinica, Qualtrics, and REDCap stand out as essential tools for managing clinical data efficiently.

- **OpenClinica** is a powerful open-source electronic data capture (EDC) system designed for clinical trials and observational studies. It supports regulatory-compliant data management in accordance with Good Clinical Practice (GCP), FDA 21 CFR Part 11, and other global standards. OpenClinica offers advanced features such as case report form (CRF) design, real-time data validation, and role-based access control. Its ability to integrate with external systems via APIs allows for seamless data exchange, making it a robust choice for large-scale clinical research. Additionally, OpenClinica provides audit trails, electronic signatures, and automated workflows, ensuring transparency and reproducibility.

- **Qualtrics** is a versatile survey platform commonly used for collecting patient-reported outcomes, quality-of-life assessments, and healthcare experience surveys. It is known for its intuitive interface, customizable survey logic, and sophisticated data analytics. In clinical research, Qualtrics is particularly useful for gathering qualitative and quantitative data from study participants, supporting both longitudinal and cross-sectional study designs. The platform includes built-in compliance with HIPAA and GDPR, making it suitable for sensitive healthcare-related surveys. Additionally, Qualtrics integrates with statistical analysis tools such as SPSS and R, facilitating seamless data processing and reporting.

- **REDCap** (Research Electronic Data Capture), my personal favorite of the three,  is a secure, web-based application designed for managing clinical and translational research data. Developed by Vanderbilt University, REDCap is widely adopted by academic and healthcare institutions for its flexibility and user-friendly interface. The platform allows researchers to create custom data collection forms, implement branching logic, and perform automated data validation. REDCap supports multi-site collaborations, ensuring secure data sharing among research teams while maintaining strict access controls. Its compliance with regulatory standards, such as HIPAA and 21 CFR Part 11, makes it a preferred choice for clinical studies that require high levels of data security. Additionally, REDCap's integration with external databases and statistical software enhances its functionality for complex clinical research projects.  Of interest to this group, REDCap has a well-established API functionality that easily allows for integration with R.  As will be discussed further in a subsequent section, there are numerous R Packages specifically built to handle REDCap data, thereby streamlining the process even further.

These tools each serve specific purposes in clinical data management, from structured data collection in clinical trials (OpenClinica) to patient-reported data collection (Qualtrics) and customizable, secure data management for diverse research needs (REDCap). By selecting the appropriate tool based on study requirements, researchers can optimize data quality, enhance efficiency, and ensure compliance with ethical and regulatory guidelines.


## Clinical Data R Tools

I wanted to highlight some of my personal favorite R Packages.I work in [REDCap](https://project-redcap.org/) often.

### REDCap-specific Packages

- ***`REDCapR`*** Facilitates secure and efficient data transfer between R and REDCap (Research Electronic Data Capture), a widely used web-based platform for managing research and clinical study data.
- ***`REDCapAPI`***Provides a flexible interface for interacting with REDCap's API, enabling users to efficiently retrieve, update, and manage data stored in REDCap databases. It offers a comprehensive set of functions for querying records, exporting metadata, importing data, handling users and project settings, and managing files or logs. Unlike REDCapR, which focuses on streamlined data extraction, REDCapAPI provides more granular control over API calls, making it well-suited for advanced users who need custom queries, fine-tuned data manipulation, and administrative project management.
- ***`redcapcustodian`*** Designed for managing user permissions and project access within REDCap using the API. It helps administrators automate tasks such as assigning roles, modifying user privileges, and auditing access logs across multiple REDCap projects.
- ***`redcapDM`***Designed to facilitate data management and quality control for research studies using REDCap. It provides functions for data cleaning, validation, monitoring, and reporting, helping researchers ensure data integrity before analysis. Key features include identifying missing or inconsistent data, summarizing project metadata, generating automated reports, and tracking changes over time.</li>
</ul>

### Non-REDCap Packages

- ***`tidygeocoder`***Simple and efficient way to perform geocoding and reverse geocoding within a tidyverse-friendly framework. It allows users to convert addresses into latitude and longitude coordinates (geocoding) and vice versa (reverse geocoding) using various online services such as OpenStreetMap (Nominatim), Google, Bing, and Census Bureau APIs. The package supports batch processing, making it useful for handling large datasets, and integrates well with dplyr, enabling seamless incorporation into data analysis workflows.
- ***`beepr`***Play notification sounds to signal the completion of code execution. It is especially useful for long-running scripts, alerting users when a task is finished without requiring them to monitor the console.
- ***`DataExplorer`***Simplifies exploratory data analysis (EDA) by providing automated functions for data visualization, summary statistics, and feature engineering. It enables users to quickly generate descriptive reports, identify missing values, visualize distributions, detect correlations, and assess data structure. Key functions include `introduce()` for dataset overview, `plot_missing()` for missing data patterns, `plot_histogram()` for distributions, and `create_report()` for generating a comprehensive EDA report.


Most people are familiar with RStudio for the purposes of having a nice user interface for interacting with R.  Although a Google Search may show that there are a number of other UIs such as `rattle` and `RKWard`, I've only found one that works on the current version of R.  Called `GrapheR`, it can be used to make simple graphing tasks easier for novice, it is not for the advanced user as the graphics are far too rudimentary.

- `GrapheR` Helps users create and visualize networks and graph-based data. It simplifies the process of constructing graphs, visualizing them with various layouts, and analyzing network structures. The package supports directed and undirected graphs, weighted edges, and offers a range of layout options for displaying graphs, such as circular, spring, and hierarchical layouts. `GrapheR` also provides utilities for network metrics like degree centrality, betweenness, and clustering, enabling users to perform network analysis efficiently. The package integrates well with tidyverse for data manipulation and supports various export formats, including PDF and PNG for easy reporting.

The final piece of the R "puzzle" I wanted to cover was a bundle of packages and a framework called the tidyverse.  




- `tidyverse` - The `tidyverse` suite of packages created by Hadley Wickham (among many countless contributors), and the [tidyverse philosophy](https://www.tidyverse.org/) centers around a consistent, human-readable approach to data science, emphasizing tidy data principles where each variable is a column, each observation is a row, and each value is a cell. It promotes a functional programming style with clear, pipeable workflows, making data manipulation, visualization, and analysis more intuitive and efficient.  Check out [the original `tidyverse` paper](https://joss.theoj.org/papers/10.21105/joss.01686).

For the purposes of this tutorial, we're going to be using the following:

- ***`sqldf`***perform SQL queries directly on R data frames, facilitating the manipulation and analysis of data using SQL syntax. It enables seamless integration of SQL operations such as SELECT, JOIN, GROUP BY, and WHERE with R data structures, without the need to convert data into a database format. `sqldf` can handle both simple and complex queries and is particularly useful for users familiar with SQL who prefer its syntax for data manipulation. The package supports working with data stored in data frames, tables, or external databases like SQLite and MySQL.

- ***`ggplot2`*** One of the most widely used libraries for creating data visualizations. It is based on the Grammar of Graphics, which provides a systematic approach to building plots by combining various components such as data, aesthetic mappings (e.g., x and y axes, color, size), geometries (e.g., points, lines, bars), and statistical transformations. With `ggplot2`, users can easily create a wide range of static and dynamic visualizations, including scatter plots, bar charts, histograms, line graphs, boxplots, and more. It also provides robust customization options for themes, labels, and scales, allowing users to tailor visualizations for publication-quality graphics. The `ggplot2` package integrates well with other tidyverse packages, making it a powerful tool for data exploration, analysis, and presentation.

- ***`survival`***comprehensive tool for performing survival analysis and modeling time-to-event data. It provides functions to fit and analyze various survival models, including the Cox proportional hazards model, Kaplan-Meier estimators, and parametric survival models (e.g., exponential, Weibull). The package includes methods for estimating survival curves, performing log-rank tests for comparing survival between groups, and handling censored data (where an event has not occurred for some individuals by the end of the study). The survival package also allows for advanced statistical techniques such as frailty models and competing risks analysis. It is widely used in clinical research, epidemiology, and other fields involving time-to-event data.

- ***`survminer`***specifically designed for visualizing survival analysis results. It provides functions to create publication-ready survival plots, such as Kaplan-Meier curves, survival curves by groups, and hazard ratio plots. The package allows for easy customization of plots, including adjustments to colors, labels, legends, and axes, and supports the addition of confidence intervals and p-values to plots. `survminer` works seamlessly with the survival package in R, which is used to perform survival analysis. It is particularly useful for researchers and analysts working with clinical or time-to-event data, enabling them to generate clear, informative, and visually appealing plots to communicate survival outcomes effectively.

To use these in R we need to first load them in memory like so:

```{r, comment=FALSE, message=FALSE}
library(sqldf)
library(ggplot2)
library(survival)
library(survminer)
```


## Lung Data
We will use the lung dataset available in the `survival` package. The data contain subjects with advanced lung cancer from the North Central Cancer Treatment Group. It includes the 10 following variables:

- inst:  Institution code
- time:  Survival time in days
- status:  censoring status 1=censored, 2=dead
- age:  Age in years
- sex:  Male=1 Female=2
- ph.ecog:  ECOG performance score as rated by the physician. 0=asymptomatic, 1= symptomatic but completely ambulatory, 2= in bed <50% of the day, 3= in bed > 50% of the day but not bedbound, 4 = bedbound
- ph.karno:  Karnofsky performance score (bad=0-good=100) rated by physician
- pat.karno:  Karnofsky performance score (0 = bad, 100 = good) as rated by patient
- meal.cal:  Calories consumed at meals
- wt.loss:  Weight loss in last six months

[source](https://rpubs.com/floreuzan/survivalanalysislungdata)


Now that we have loaded the `survival` package, we can check it out just by typing `lung`. We will use the `head()` function to just preview the first several rows of data.  

```{r}
head(lung)
```


## SQL for Data Management

The `sqldf` package allows users to execute SQL queries on R data frames, providing a convenient way to subset, filter, and summarize data. It is especially useful for those familiar with SQL but not as comfortable with R’s native data manipulation functions like `dplyr` or `data.table`. 

Since SQL is often considered more intuitive and readable, `sqldf` enables users to write queries in a "plain English" style rather than learning complex R-specific syntax. One of its advantages is that it treats R data frames as database tables, allowing users to perform operations like filtering, joining, grouping, and ordering without converting the data into another format. The queries are executed using an in-memory SQLite database by default, ensuring efficiency without the need for an external database system. The basic syntax follows standard SQL conventions, with the query enclosed in double quotes within the `sqldf()` function.

For example, if you wanted to retrieve all columns for the first 10 rows in the `lung` dataset where age > 60, you would use:
`sqldf("select * from lung where age>60 limit 10")`

```{r lung1, echo=FALSE}
sqldf("select * from lung where age>60 limit 10")
```

The package supports a variety of operations, making it a versatile tool for data analysis.

- It allows for easy data filtering and selection, enabling users to extract specific rows and columns with simple queries. 
- Aggregation and summarization functions, such as `COUNT`, `AVG`, `MIN`, and `MAX`, allow users to quickly compute statistics on datasets. 
- Sorting and ordering data is straightforward with the `ORDER BY` clause, and merging datasets is effortless using SQL joins, such as `INNER JOIN` and `LEFT JOIN`. 
- Additionally, `sqldf` can be used for data cleaning and validation, helping identify duplicates, missing values, or inconsistencies within a dataset.

Here is a basic example of using `sqldf` to create a basic summary table: 


```{r, eval = FALSE}
sqldf("select sex, 
      count(*),  
      count(case when (age<=60) then 1 end) as LTE60, 
      count(case when (age>60) then 1 end) as GT60 
      from lung group by sex order by sex")
```

Here's the output:<br>
```{r lung3, echo = FALSE}
sqldf("select sex, 
      count(*),  
      count(case when (age<=60) then 1 end) as LTE60, 
      count(case when (age>60) then 1 end) as GT60 
      from lung group by sex order by sex")
```

This is a good example of how a more advanced query is still easy to understand, allowing for even complex SQL queries to be relatively easy to understand by new users. 

One of the biggest benefits of sqldf is its familiar syntax, making it an ideal choice for users who are comfortable with SQL but not R’s built-in data manipulation tools. It is lightweight and efficient, as it uses an in-memory SQLite database for quick execution, and requires no additional database setup. Since it works directly with R data frames, there is no need to export data into a separate SQL environment.

However, while `sqldf` is excellent for quick exploratory analysis and ad-hoc querying, it **may not be the best choice for very large datasets**, where optimized R packages like `dplyr` or `data.table` provide faster performance. Nonetheless, for small- to medium-sized datasets, `sqldf` remains an accessible, powerful, and intuitive tool for SQL-based data analysis in R.

## Exploring `ggplot2`

`ggplot2` is a complete framework for developing complex graphs, with an extensive series of options and customizations available. Based on <a href="https://www.amazon.com/Grammar-Graphics-Leland-Wilkinson/dp/0387245448">The Grammar of Graphics by Leland Wilkinson</a>, it provides a structured approach to creating a wide variety of static, animated, and interactive visualizations with a consistent syntax. One of its key features is the layered approach, where plots are built step by step by adding components like points, lines, bars, and annotations. The aesthetic mapping `(aes())` function allows users to define how data variables correspond to visual properties such as axis positions, colors, sizes, and shapes.

Different types of plots are created using geometries `(geom_xxx)` including:

- `geom_point()` for scatter plots
- `geom_line()` for line charts
- `geom_bar()` for bar plots
- `geom_histogram()` for histograms

For multi-panel displays, faceting allows splitting data into smaller subplots using `facet_wrap()` or `facet_grid()`. 

The package also provides themes `(theme_xxx)` for customizing plot appearance, including background, text size, legend placement, and grid lines. 

Users can further control plot elements through scales `(scale_*)`, which adjust axis limits, color gradients, and transformations.

Incorporating statistical summaries is easy with statistical transformations `(stat_*)`, such as `geom_smooth()` for regression lines and `geom_density()` for density estimates. Additionally, extensions like `gganimate` for animations, ggthemes for additional styles, and `patchwork` for arranging multiple plots enhance its functionality. 

The `ggplot2` package  is widely favored for its declarative syntax, which separates data, aesthetics, and layers, making complex visualizations intuitive and reproducible. It is highly flexible and extensible, making it an essential tool for data visualization in R.

To create a basic scatterplot using the `lung` data, here's how I would do it:

```{r, eval=FALSE}
ggplot(lung, aes(x = age, y = time, color = factor(sex))) +
  geom_point(size = 3, alpha = 0.7) +  # Points with transparency
  labs(
    title = "Survival Time vs. Age",
    x = "Age (years)",
    y = "Survival Time (days)",
    color = "Sex"
  ) +
  scale_color_manual(values = c("blue", "red"), labels = c("Male", "Female")) +
  theme_minimal()
```

Here's the output:<br>
```{r plotsample, echo=FALSE}
ggplot(lung, aes(x = age, y = time, color = factor(sex))) +
  geom_point(size = 3, alpha = 0.7) +  # Points with transparency
  labs(
    title = "Survival Time vs. Age",
    x = "Age (years)",
    y = "Survival Time (days)",
    color = "Sex"
  ) +
  scale_color_manual(values = c("blue", "red"), labels = c("Male", "Female")) +
  theme_minimal()
```

- `ggplot(lung, aes(x = age, y = time, color = factor(sex)))` Defines the dataset and aesthetic mappings (age on the x-axis, survival time on the y-axis, and color based on sex).
- `geom_point(size = 3, alpha = 0.7)` Creates a scatter plot with medium-sized points and slight transparency for better visibility.
- `labs(...)` Customizes the title, axis labels, and legend title.
- `scale_color_manual(...)` Manually assigns colors (blue for males and red for females) and renames legend labels.
- `theme_minimal()` Uses a clean, minimalistic theme for a better visual appearance.

## Introduction to Kaplan-Meier Curves

KM Curves are also known as Survival Plots or Time To Event Analyses. I first started learning about them when I was the data analyst for the Solid Organ Transplant program at SickKids Hospital, where I had to report on the time to Death, Re-Transplant or Transplant of a Different Organ.  Using KM Curves, I was able to easily break down the various organ groups into different categories and explore the relationships that helped drive the decisions in the program.  

Kaplan-Meier curves are a statistical method used to estimate and visualize survival probabilities over time, commonly applied in medical research and clinical studies. These curves illustrate the proportion of individuals who have not yet experienced a specific event, such as death, disease recurrence, or recovery, at various time points. The curve starts at 100%, representing all individuals in the study, and steps down each time an event occurs. Unlike simple averages, Kaplan-Meier analysis accounts for "censored" data, which includes participants who are lost to follow-up or do not experience the event before the study ends. These individuals are still included in calculations up to the point they leave the study but do not contribute beyond that.

Kaplan-Meier curves are particularly useful for comparing survival outcomes between different groups, such as patients receiving different treatments. If one group's curve declines more steeply than another’s, it suggests a higher event rate and worse outcomes, while a flatter curve indicates better survival or a lower risk of the event occurring. The statistical significance of differences between curves is typically assessed using the log-rank test, which evaluates whether survival distributions differ meaningfully between groups. Researchers often use these curves to make informed decisions about treatment efficacy, disease progression, and risk factors. By providing a clear, time-based visualization of survival probabilities, Kaplan-Meier curves help guide clinical decisions and shape future research directions.

## `survival` Introduction

The `survival` R package is a widely used tool for analyzing time-to-event data, also known as survival data, which is commonly applied in fields like clinical research, epidemiology, and social sciences. Survival analysis focuses on estimating the time it takes for a specific event, such as death, disease recurrence, or recovery, to occur, and it handles censored data—cases where the event has not yet occurred by the study's end. 

The package provides a range of statistical methods including:

- The **Kaplan-Meier estimator** which is used to visualize survival functions and estimate the probability of an event occurring over time. 
- The **Cox proportional hazards model**, a powerful method for examining the effect of multiple covariates (like age, treatment, or other risk factors) on the hazard or risk of an event happening. 
- **Log-rank tests** to compare survival distributions between groups, helping researchers determine if there are significant differences in survival between different treatments or population groups. 
- parametric survival models such as exponential, Weibull, or log-normal can be applied when the assumptions of the Cox model do not hold, offering more flexibility in modeling time-to-event data. 

The survival package also supports the analysis of competing risks, where individuals may experience one of several possible events, and the occurrence of one event precludes the occurrence of others. Researchers can also visualize and compare survival curves for different groups, further aiding in understanding the impact of various factors on survival outcomes. With its comprehensive set of tools for modeling and visualizing survival data, the survival package is essential for researchers aiming to identify factors that influence the timing of events and make informed decisions based on time-based data.

## `survival` Example


``` {r lung5, eval =FALSE}
# Create a survival object: Surv(time, status)
#'time' is survival time
#'status' is event (1 = death, 0 = censored)
surv_obj <- Surv(lung$time, lung$status)

# Fit the Kaplan-Meier model
km_fit <- survfit(surv_obj ~ 1) 
# '~ 1' indicates no grouping variable, just the overall survival curve

# Plot the Kaplan-Meier survival curve
plot(km_fit, main="Kaplan-Meier Survival Curve", 
     xlab="Time (days)", ylab="Survival Probability")
```

Here is the output: 
<br> 
``` {r lung5_out, echo = FALSE}
# Create a survival object: Surv(time, status)
#'time' is survival time
#'status' is event (1 = death, 0 = censored)
surv_obj <- Surv(lung$time, lung$status)

# Fit the Kaplan-Meier model
km_fit <- survfit(surv_obj ~ 1) 
# '~ 1' indicates no grouping variable, just the overall survival curve

# Plot the Kaplan-Meier survival curve
plot(km_fit, main="Kaplan-Meier Survival Curve", 
     xlab="Time (days)", ylab="Survival Probability")
```

### `survival` Example - Explanation


``` {r lung5_again, eval =FALSE}
# Create a survival object: Surv(time, status)
#'time' is survival time
#'status' is event (1 = death, 0 = censored)
surv_obj <- Surv(lung$time, lung$status)

# Fit the Kaplan-Meier model
km_fit <- survfit(surv_obj ~ 1) 
# '~ 1' indicates no grouping variable, just the overall survival curve

# Plot the Kaplan-Meier survival curve
plot(km_fit, main="Kaplan-Meier Survival Curve", 
     xlab="Time (days)", ylab="Survival Probability")
```

- `surv_obj <- Surv(lung$time, lung$status)`Creates a survival object using the `Surv()` function, which combines survival time `(lung$time)` and event status `(lung$status)`. Here, the event status is coded as 1 for an event (e.g., death) and 0 for censored data, meaning the event did not occur before the study ended.

- `km_fit <- survfit(surv_obj ~ 1)` The Kaplan-Meier model is fitted using the `survfit()` function. The formula `surv_obj ~ 1` indicates that the model estimates an overall survival curve without considering any grouping variables.

- `plot(km_fit, main="Kaplan-Meier Survival Curve", xlab="Time (days)", ylab="Survival Probability")` The plot() function generates a Kaplan-Meier survival curve, visually displaying how survival probability changes over time. The main argument sets the title of the plot, while `xlab` and `ylab` label the x-axis `Time (days)` and y-axis `Survival Probability`, respectively.

The Kaplan-Meier survival curve generated from this code provides a simple overview of survival probabilities over time for the entire dataset, without breaking it down by groups. The x-axis represents time in days, while the y-axis shows the probability of survival. The curve starts at 1 (100% survival) and gradually steps downward as events (e.g., deaths) occur. Each drop in the curve corresponds to a time point where at least one event happened. The presence of censored data (patients lost to follow-up or still alive at the study's end) is typically indicated by small marks on the curve.

While this plot gives a good initial look at overall survival trends, it lacks detail on differences between subgroups. For example, it doesn’t show whether survival differs based on factors like **gender**, **treatment**, or **disease severity**. To gain deeper insights, grouped survival curves and statistical comparisons (like the log-rank test) would be needed.

### `survival` Example 2

Suppose we want to compare survival between two groups (e.g., based on the variable `lung$sex`).

``` {r lung6, echo = FALSE}

# Fit Kaplan-Meier models for each group
km_fit_group <- survfit(surv_obj ~ lung$sex)

# Plot the survival curves by gender
plot(km_fit_group, col=c("blue", "red"), lty=c(1, 2),
     main="Survival by Gender", 
     xlab="Time (days)", 
     ylab="Survival Probability")

legend("topright", legend=c("Male", "Female"), col=c("blue", "red"), lty=c(1, 2))

# Perform a log-rank test to compare survival between males and females
log_rank_test <- survdiff(surv_obj ~ lung$sex)
log_rank_test
```

## `survival` Example 2 Explanation

So how do we make the above plot? We can now also include a variable to fit for different groups based on that variable. 

``` {r lung6_echo, eval = FALSE}

# Fit Kaplan-Meier models for each group
km_fit_group <- survfit(surv_obj ~ lung$sex)

# Plot the survival curves by gender
plot(km_fit_group, col=c("blue", "red"), lty=c(1, 2),
     main="Survival by Gender", 
     xlab="Time (days)", 
     ylab="Survival Probability")

legend("topright", legend=c("Male", "Female"), col=c("blue", "red"), lty=c(1, 2))

# Perform a log-rank test to compare survival between males and females
log_rank_test <- survdiff(surv_obj ~ lung$sex)
log_rank_test
```


-`km_fit_group <- survfit(surv_obj ~ lung$sex)` This code builds upon the Kaplan-Meier survival model we previously examined by incorporating gender as a grouping variable. Instead of generating a single survival curve for the entire dataset, the model now estimates separate survival curves for males and females using the formula `surv_obj ~ lung$sex`. This allows for a direct visual and statistical comparison of survival outcomes between the two groups.

- Plotting the data:
```
plot(km_fit_group, col=c("blue", "red"), lty=c(1, 2), main="Survival by Gender", 
xlab="Time (days)", ylab="Survival Probability")
legend("topright", legend=c("Male", "Female"), col=c("blue", "red"), lty=c(1, 2))
```
The next step is to plot these survival curves using the plot() function. In this version, the plot assigns blue to males and red to females, with line styles differentiated—a solid line for males and a dashed line for females. The x-axis represents time in days, while the y-axis shows the probability of survival over time. A legend is added to the top right corner to help distinguish between the two survival curves.

-```log_rank_test <- survdiff(surv_obj ~ lung$sex) log_rank_test``` 
To formally test whether survival differs between males and females, the `survdiff()` function performs a log-rank test, a statistical method used to compare survival distributions between groups. The output includes a chi-square statistic and a p-value, where a p-value below 0.05 would indicate a statistically significant difference in survival between the two genders. This approach provides more meaningful insights than the earlier single-curve model by highlighting potential disparities in survival outcomes between patient groups.


## `survminer` Introduction

As you may have noticed, the `survival` outputs are OK but not something you'd want to use in a presentation.  This is where the `survminer` package comes in - it's a powerful tool designed to help users visualize the results of survival analysis in a more informative and publication-ready format. It works seamlessly with the `survival` package, which is used for fitting survival models, to generate a variety of plots that aid in the interpretation of time-to-event data. One of the primary features of `survminer` is its ability to create Kaplan-Meier curves, which estimate and display survival probabilities over time. These curves can be further customized by adding p-values, confidence intervals, and other statistical details, making it easier to interpret the significance of survival differences between groups.

In addition to Kaplan-Meier curves, `survminer` allows for the visualization of survival curves for multiple groups, making it especially useful for comparing the survival outcomes of different treatment groups or patient subpopulations. The package also includes options to visualize hazard ratios and other important statistical measures, which are crucial for understanding the impact of covariates on survival. The graphical tools provided by `survminer` support clear, attractive, and customizable plots that can be tailored to meet specific research needs, from adjusting colors and labels to modifying plot legends and axes.

Furthermore, `survminer` offers additional visualization techniques, such as:

- Survival curves by strata
- Cox regression plots
- Diagnostic plots for assessing the fit and assumptions of survival models

These plots help users identify potential issues, such as proportional hazards violations, and gain deeper insights into the factors influencing survival outcomes.

By integrating easily with the `survival` package, `survminer` makes it much simpler to communicate complex survival analysis results visually. It is widely used in clinical research, epidemiology, and other fields where time-to-event data plays a key role, providing researchers with effective tools to present and interpret survival data in a clear, accessible way. With its intuitive interface and powerful visualization capabilities, `survminer` is an invaluable resource for anyone working with survival data.

`survminer` is an R package designed to simplify the visualization of survival analysis results, particularly those generated using the survival package. It provides a user-friendly and elegant way to create Kaplan-Meier plots, Cox proportional hazards models, and other survival-related visualizations. The package is widely used in clinical and epidemiological research, where understanding time-to-event data is crucial.   

As it is dependent on the installation of `ggplot2`, it uses a lot of `gg` styled syntax, including one of the key functions `ggsurvplot()`, which creates Kaplan-Meier survival curves with extensive customization options. Users can modify axis labels, legends, colors, confidence intervals, risk tables, and p-values with ease. The function integrates well with `ggplot2`, allowing additional modifications through layers and themes. Other functions like `ggadjustedcurves()` enable the visualization of adjusted survival curves based on Cox models, while `ggforest()` generates forest plots for hazard ratios.

`Survminer` also offers tools for assessing proportional hazards assumptions using `ggcoxzph()`, which plots scaled Schoenfeld residuals[^1] to check for violations. Additionally, `ggcompetingrisks()` helps visualize competing risks in survival data. These visualization tools help researchers interpret survival models more effectively and present results in a publication-ready format.

[^1]: Schoenfeld residuals are a type of diagnostic tool used to assess the proportional hazards assumption in Cox regression models, helping to determine whether the hazard ratio between groups remains constant over time.

## `survminer` Example
Here is a sample KM Curve using the same data as the previous examples above: 

```{r survival, echo=FALSE}
fit <- survfit(Surv(time, status) ~ sex, data = lung)

ggsurvplot(fit, data = lung, 
           pval = TRUE, 
           risk.table = TRUE, 
           conf.int = TRUE, 
           ggtheme = theme_minimal())
```

### `survminer` Example - Explanation 

The following code makes the above plot:

```{r survival_echo, eval = FALSE}
fit <- survfit(Surv(time, status) ~ sex, data = lung)

ggsurvplot(fit, data = lung, 
           pval = TRUE, 
           risk.table = TRUE, 
           conf.int = TRUE, 
           ggtheme = theme_minimal())
```


- `fit <- survfit(Surv(time, status) ~ sex, data = lung)`
This line fits a Kaplan-Meier survival model using the `survfit()` function. It defines the time-to-event variable (time) and the event status (status) using the `Surv()` function, while sex is included as a grouping variable. This means the model will estimate and compare survival curves separately for males and females using data from the lung dataset.

- `ggsurvplot(fit, data = lung, pval = TRUE, risk.table = TRUE, conf.int = TRUE, ggtheme = theme_minimal())`
  - The `ggsurvplot()` function is used to create an enhanced survival plot. 
  - The first argument, `fit`, is the Kaplan-Meier model object, and `data = lung` specifies the dataset. The `pval = TRUE` argument adds a p-value to the plot, indicating whether there is a significant difference in survival between the two groups.  A p-value less than 0.05 suggests that the differences in survival between these groups are statistically significant. 
  - The `risk.table = TRUE` option includes a risk table below the plot, showing the number of individuals still at risk at different time points. 
  - The `conf.int = TRUE` argument adds confidence intervals, which provide a shaded region around the survival curves to indicate uncertainty. 
  - Finally, `ggtheme = theme_minimal()` applies a clean, minimalistic theme to improve the plot’s readability and aesthetics.

The resulting Kaplan-Meier plot gives a clear visual representation of the survival probabilities over time for different groups. By examining the curves, you can quickly determine whether there is a notable difference in survival between males and females. If the curves diverge significantly, this may indicate that sex plays a role in the survival time of lung cancer patients. The confidence intervals help assess the precision of the survival estimates, while the risk table provides more detailed information on how many individuals are still at risk at each time point. Overall, the Kaplan-Meier curve, along with the statistical tests and additional plot elements, serves as an invaluable tool for understanding survival patterns and comparing groups in clinical studies.

## KM Curve Interpretation

It is not necessarily true that if two Kaplan-Meier (KM) curves cross, there is no statistical significance between the groups. The crossing of KM curves does not automatically imply the absence of significance. In fact, the interpretation of crossing curves is more nuanced and requires further analysis.

If two KM curves cross, it suggests that at different time points, one group may have a higher survival probability than the other, but this relationship changes over time. This could be due to differences in how the groups behave at different stages of the study. The crossing of the curves indicates that the survival advantages or disadvantages between the groups are not consistent across time, which could reflect a change in the factors affecting survival as the study progresses.

The log-rank test, which is often used to compare two or more survival curves, tests the overall difference in survival between groups. The test evaluates whether the curves are significantly different across the entire time period, even if they cross at certain points. A significant log-rank test (p < 0.05) means that there is a difference in survival between the groups, regardless of whether the curves cross. The test considers the entire survival distribution and is not affected by the crossing of the curves at specific time points.

However, crossing curves can complicate interpretation, especially if the curves cross multiple times. In such cases, the proportional hazards assumption (which underpins the Cox proportional hazards model) may be violated. This assumption suggests that the hazard ratio between the groups is constant over time, but with crossing curves, the relationship between the groups may change at different intervals. This indicates that the survival patterns of the groups are not proportional over time, and more complex analysis might be needed to fully understand the relationship.

Even when KM curves cross, it is still important to consider the clinical context and look at other statistical measures. For instance, median survival times, confidence intervals, and hazard ratios provide additional insights into the significance of the survival differences between groups. These metrics help provide a clearer understanding of the survival trends, even in the presence of crossing curves.

In conclusion, while crossing KM curves can indicate that the survival patterns of two groups differ over time, it does not imply that there is no significance. A log-rank test or other statistical methods can still show whether there is a significant difference in survival between the groups across the entire study period, even in the presence of crossing curves. Therefore, the crossing of the curves should be interpreted in conjunction with statistical tests and clinical context.

## Putting it all together

Having explored various methods for generating Kaplan-Meier (KM) curves, we can now outline a hypothetical workflow that demonstrates the use cases for each of the packages we’ve discussed. This workflow will take us from data extraction to survival analysis and visualization, showcasing how different tools contribute to the process.

Data Extraction and Filtering:
We begin by leveraging an SQL query to extract and filter our dataset based on specific conditions. For instance, we may select patients within a defined age range, those diagnosed with diabetes, and individuals residing in a particular geographic region. This step ensures that our analysis is conducted on a well-defined and relevant subset of data.

Survival Model Construction:
With our filtered dataset, we proceed by creating a `surv` object, which serves as the foundation for our survival analysis. We then fit a Cox proportional hazards regression model to evaluate the impact of key variables—such as sex and ph.ecog (a performance status score)—on survival outcomes. To gain deeper insights, we conduct this modeling process separately for each of the three data subsets, allowing for a more nuanced comparison across different patient groups.

Kaplan-Meier Curve Visualization:
Next, we utilize the `survminer` package to generate Kaplan-Meier survival curves, stratified by gender. By visualizing survival probabilities separately for male and female patients within each subset, we can identify potential differences in survival trends and explore patterns that might inform further investigation.

Statistical Comparison of Survival Distributions:
Finally, we conduct log-rank tests to statistically compare survival distributions between males and females. This test assesses whether there is a significant difference in survival probabilities between the two groups, providing a formal statistical basis for evaluating gender-based disparities in survival outcomes.

Through this workflow, we systematically apply SQL for data extraction, survival analysis techniques for modeling, and visualization tools for interpretation—demonstrating the utility of each package in the survival analysis pipeline.




Hopefully this will give you a sense of how you can quickly pull together complex data and generate powerful, useful visualizations and analyses. 

## Conclusion
R is an essential tool in modern research analysis due to its versatility, powerful statistical capabilities, and large community of users who contribute to its continuous development. It is widely used across diverse fields, including medicine, social sciences, economics, and engineering, because it can handle a broad range of data types and offers specialized packages for various statistical analyses. R allows researchers to not only perform complex statistical tests but also visualize data effectively, making it a comprehensive choice for data analysis and presentation. One of the standout features of R is its open-source nature, making it accessible to anyone with an interest in data analysis without the need for expensive software licenses. Additionally, R’s extensive range of packages, such as `ggplot2` for visualization, `survival` for survival analysis, and `dplyr` for data manipulation, means that researchers can easily apply sophisticated methods without reinventing the wheel. R is also highly reproducible, which is vital for transparency in research, enabling others to replicate analyses and confirm findings.

A particularly important statistical method available in R is the Kaplan-Meier (KM) curve, which is invaluable in survival analysis. KM curves are crucial for visualizing and interpreting time-to-event data, especially in clinical research where understanding patient survival is key. By displaying the probability of survival over time for different groups, KM curves provide insight into how factors such as treatment, age, or sex may influence outcomes. They are widely used in clinical trials to compare the effectiveness of different treatments, and in epidemiological studies to understand the impact of various risk factors on survival. The ability to analyze and interpret survival data using KM curves in R has made the software indispensable in both clinical research and public health. The log-rank test, commonly used alongside KM curves, allows researchers to statistically determine whether differences in survival across groups are significant, providing a comprehensive approach to survival analysis.

Despite its power, R can initially be intimidating, especially for beginners who are unfamiliar with programming or statistical analysis. The syntax, the wide variety of functions, and the multitude of available packages can seem overwhelming. However, once the basics are learned, R becomes much more approachable. Key concepts such as understanding data structures (like vectors, data frames, and lists), mastering basic operations, and knowing how to use simple functions are the foundation for using R effectively. Once these fundamentals are grasped, learning more advanced techniques, such as performing specific statistical tests or generating complex visualizations, becomes much easier. Furthermore, the supportive R community and abundant online resources, including tutorials, forums, and documentation, make it easier to troubleshoot and learn at your own pace. As a result, while R may appear daunting at first, it becomes a highly rewarding tool once the initial learning curve is overcome. With practice, users can harness its full potential, making R an essential asset for any researcher.

In conclusion, R is an indispensable tool for researchers due to its flexibility, statistical power, and ease of reproducibility. Kaplan-Meier curves exemplify the power of R in survival analysis, providing essential insights into time-to-event data. While R may seem intimidating initially, learning the basics unlocks a wealth of possibilities for data analysis and visualization, making it an invaluable skill for any researcher.

